---
title: "Final Project FA"
author: "Can Dai"
date: '2022-08-11'
output: pdf_document
---

## Problem 1

Choose a technique that we have covered so far in this course, and try applying that technique to your data. You may choose any of
a) Model building and Multiple Regression b) PCA c) CFA d) CCA e) CA (correspondence analysis)
If you are working as a group, each member of your group should try a different technique, or the same technique with different aspects of the data.

```{r}
############## Common Factor Analysis for Patient Survival Dataset #############
# Dataset cleaning

library(dplyr)
# Factor Analysis on Patient Survival Dataset
dataset = read.csv(file="dataset.csv", header=TRUE, sep=",")

dim(dataset)
str(dataset)
#names(dataset)

#Remove ID columns
datasetv1 = dataset[,-c(1,2,3,13)]
#str(datasetv1)

datasetv2 = datasetv1[,which(colMeans(!is.na(datasetv1))>0.5)]
sum(is.na(datasetv2))


NACheck = lapply(datasetv1, is.na)
NACheckSum = lapply(NACheck, sum)
cbind(NACheckSum)

NACheckv2 = lapply(datasetv2, is.na)
#NACheckv2
NACheckSumv2 = lapply(NACheckv2, sum)
#NACheckSumv2
cbind(NACheckSumv2)

#Listwise Deletion for variables with less than 50% of NAs in their rows. 
datasetv3 <- na.omit(datasetv2)
#dim(datasetv3)

sum(is.na(datasetv3))

#str(datasetv3)

datasetv4 <- datasetv3 %>% dplyr::select(where(~!is.character(.))) %>% glimpse()

#str(datasetv4)

### Remove binary columns except target column (hospital_death)

datasetv5 <- datasetv4 %>% dplyr::select(-hospital_death, -elective_surgery,-readmission_status,-apache_post_operative,-arf_apache,
                                  -gcs_unable_apache,-intubated_apache,-ventilated_apache,-aids,
                                  -cirrhosis,-diabetes_mellitus,-hepatic_failure,-immunosuppression,-leukemia,
                                  -lymphoma,-solid_tumor_with_metastasis) %>% glimpse()
#str(datasetv5)
#dim(datasetv5)

```

```{r}
#### Remove non continuous numerical variables

datasetv6 <- datasetv5[,-c(6,7,10,11,12)]

str(datasetv6)
dim(datasetv6)
head(datasetv6)


```

```{r warning=FALSE}
# Factoribility / Reliability Tests

#Test KMO Sampling Adequacy
library(psych)
KMO(datasetv6)
#Overall MSA =  0.81

#Test Bartlett's Test of Sphericity => checking for correlation 
library(REdaS)
bart_spher(datasetv6)
#p-value < 2.22e-16 (Very Small Number)

#Test for Reliability Analysis using Cronbach's Alpha => Looking for >0.7 => these data points belong together
alpha(datasetv6,check.keys=TRUE)
#raw_alpha = 0.8199


```

KMO value is 0.81, the Bartlett's Test of Sphericity indicate that the p-value < 2.22e-16 and the Cronbach's Alpha test indicate that raw_alpha = 0.8199. These values indicate that the data is reliable and stable.

```{r}
# Create a Scree plot to see the number of factors to use
p = prcomp(datasetv6, center=T, scale=T)
p

#Check Scree Plot
plot(p)
abline(1, 0)
summary(p)
#options(max.print = 10000)  
#print(p)
table(p$sdev>1)
```

Using the scree plot above, I have decided to use the knee method to determine the number of factors to use for the FA analysis. As a result, I decided to use 5 factors since there is an elbow between 5-6 factors.

```{r warning=FALSE}
library(psych)
#Conducting Factor Analysis
fit = psych::fa(datasetv6, rotate="varimax", nfactors=5, scores=TRUE)
print(fit$loadings, cutoff=.4, sort=T)
summary(fit)
```

The cumulative variance is 0.424. The root mean square of the residuals (RMSA) is  0.08. In addition, FA was able to create 5 distinct factors out of the variables. 

The plots below show the variable loadings for each respecting factor who are bigger than 0.4:

```{r}
library(dplyr)
library(ggplot2)
fa_factors <- as.data.frame(unclass(fit$loadings))
fa_factor1_loading <- as.data.frame(fit$loadings[,1])
fa_factor1_loading_big_point4 <- fa_factor1_loading %>% filter_at(vars(1:1), any_vars(abs(.) >0.4)) 

fa_factor5_loading <- as.data.frame(fit$loadings[,2])
fa_factor5_loading_big_point4 <- fa_factor5_loading %>% filter_at(vars(1:1), any_vars(abs(.) >0.4)) 

fa_factor2_loading <- as.data.frame(fit$loadings[,3])
fa_factor2_loading_big_point4 <- fa_factor2_loading %>% filter_at(vars(1:1), any_vars(abs(.) >0.4)) 

fa_factor4_loading <- as.data.frame(fit$loadings[,4])
fa_factor4_loading_big_point4 <- fa_factor4_loading %>% filter_at(vars(1:1), any_vars(abs(.) >0.4)) 

fa_factor3_loading <- as.data.frame(fit$loadings[,5])
fa_factor3_loading_big_point4 <- fa_factor3_loading %>% filter_at(vars(1:1), any_vars(abs(.) >0.4)) 

#horizontal bar plot Factor 1
ggplot(fa_factor1_loading_big_point4, aes(x=rownames(fa_factor1_loading_big_point4), y=fa_factor1_loading_big_point4[,1]))+
  geom_bar(stat = "identity") + coord_flip() + ggtitle("Factor 1 loadings bigger than 0.4") +
  xlab("Variables") + ylab("Loadings")

#horizontal bar plot Factor 2
ggplot(fa_factor2_loading_big_point4, aes(x=rownames(fa_factor2_loading_big_point4), y=fa_factor2_loading_big_point4[,1]))+
  geom_bar(stat = "identity") + coord_flip() + ggtitle("Factor 2 loadings bigger than 0.4") +
  xlab("Variables") + ylab("Loadings")

#horizontal bar plot Factor 3
ggplot(fa_factor3_loading_big_point4, aes(x=rownames(fa_factor3_loading_big_point4), y=fa_factor3_loading_big_point4[,1]))+
  geom_bar(stat = "identity") + coord_flip() + ggtitle("Factor 3 loadings bigger than 0.4") +
  xlab("Variables") + ylab("Loadings")
#horizontal bar plot Factor 4

ggplot(fa_factor4_loading_big_point4, aes(x=rownames(fa_factor4_loading_big_point4), y=fa_factor4_loading_big_point4[,1]))+
  geom_bar(stat = "identity") + coord_flip() + ggtitle("Factor 4 loadings bigger than 0.4") +
  xlab("Variables") + ylab("Loadings")

#horizontal bar plot Factor 5
ggplot(fa_factor5_loading_big_point4, aes(x=rownames(fa_factor5_loading_big_point4), y=fa_factor5_loading_big_point4[,1]))+
  geom_bar(stat = "identity") + coord_flip() + ggtitle("Factor 5 loadings bigger than 0.4") +
  xlab("Variables") + ylab("Loadings")

```

Looking at the plots, we can describe the variables accordingly:

Factor 1: Loadings show that this factor is about a patient's highest blood pressure during their unit stay (during their first hour or during their first 24 hours).

Factor 2: Loading shows that this factor is about a patients blood urea nitrogen, creatinine, potassium concentrations.

Factor 3: Loadings indicate that this factor is about a patients heartrate, resperatition and temperature during their first hour or first 24 hours of the stay. These variables are vitals.

Factor 4: Loadings indicate that this factor is about a patients red blood cells proportion and  hemaglobin levels during their first 24 hours of stay.

Factor 5: Factor five is loadings are almost the opposite of Factor 1 and is about a patient's lowest blood pressure during their unit stay (during their first hour or during their first 24 hours).



```{r} 
# #SAMPLE PLOTS (control+shift+c)
# plot(fit$loadings[,1], 
#      fit$loadings[,2], 
#      xlim = c(-0.2, 1), 
#      ylim = c(-0.5, 1),
#      xlab = "Factor 1", 
#      ylab = "Factor 2", 
#      main = "Factor 1 vs Factor 2")
# text(fit$loadings[,1]-0.02, 
#      fit$loadings[,2]+0.03,
#       colnames(datasetv6),
#       col="blue")
# abline(h = 0, v = 0)
# 
# 
# #SAMPLE
# library(ggrepel)
# ggplot(fa_factors, aes(MR1, MR2, label = rownames(fa_factors))) +
#   geom_text_repel() +
#   geom_point(color = 'red') +
#   theme_classic(base_size = 10)
```


